---
title: "Medicago experiment"
author: "Ryan Farquharson"
date: "3 October 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Aims
This experiment was set up to:

1) compare the nodulation and nitrogen fixation of a herbicide tolerant medicago variety to the near isogenic herbicide susceptable strain in the absence and presence of chlorsulfuron - a group B herbicide; and 
2) provide mechanistic insights into how group B herbicides affect nitrogen fixation.

Different stakeholders should be able to answer different questions from the results of this experiment.

Farmers and agronomists:
I have used Group B herbicdes in my paddocks.  Is nitrogen fixation by legumes being impacted upon?
How well does this new variety, FEH-1 (Angel), grow and fix nitrogen in the presence and absence of herbicide?
How can I tell what is happening in my paddocks?
  Do I have residual group B herbicides?
  Will my plants nodulate?
  If they nodulate, will they still fix nitrogen?

Scientists:
What are the mechanisms by which group B herbicides impact on nitrogen fixation?
  Plant only
    Root growth
    Root hairs
  Rhizobia only
    Saprophytic growth
  Plant and rhizobia
    Signalling and nodule intitiation
    Nodule development
    Mutual exchange of compounds
    
## Response variables

```{r}
str(raw3)
```

Since this experiment was set up with multiple objectives, a range of response variables were measured:
  Biomass of shoots, roots and nodules
  Nitrogen contents of shoots and roots plus nodules
  Pink nodule number
  Pink nodule mass
  Non-pink nodule number
  Non-pink nodule mass
  Acetylene reduction activity

From these response variables, a number of derived variables can be calculated:
  Total biomass and nitrogen
  Nodulation (nodule number or mass on a root biomass basis (pink or total))
  Nodule actiivty (acetylene reduction activity per unit nodule mass or per nodule (pink or total))
  Nitrogen concentrations
  Amount of nitrogen fixed

The response variables chosen for analysis will need to be chosen depending on the specific questions that need to be answered.

## Experimental factors

The experiment can be considered to be factorial with 3 factors and was blocked.

Variety (2 levels):  
  Medicago truncatula cv Herald;
  Medicago truncatula cv FEH-1.

Herbicide application (2 levels): 
  Herbicide - chlorsulfuron applied to 17 day old seedlings; 
  No herbicide applied.

Rhizobia (5 levels): uninoculated without nitrogen; uninoculated with nitrogen; RRI128; RRI128 plus herbicide; RRI128 plus herbicide and washed

Block (5 blocks):  
  The experiment was set up, arranged in the glasshouse, and analysed in blocks numberred 1 to 5. Each block contained all treatments, and all pots were randomised within each block.

## How many experimental conditions?

The experiment was set up with multiple objectives, thus subsets of the data should be analysed depending on the specific research question to be answered.

The Herbicide and Variety factors can be considered as a 2 by 2 factorial (4 combinations)

Depending on which repsonse variable is being analysed, a subset of rhizobia treatments may be considered.  
For example, it makes no sense to include the uninoculated treatments in any analysis of nodulation or nodule activity.


The Block factor (5 blocks) can be considered to be uniform across all other treatments since all pots were randomised within a block.  Hence block can be considered to be an additive factor.


## Questions for Terry

ELegant ways to subset (levels)

Transformation - e.g. ara

Derived or calculated variables e.g. amount of nitrogen fixed, fixed per nodule, fixed per unit nodule mass

Nested designs - e.g. innoculated/uninoculated +-N, innoculated +H, -H, washed




