---
title: "Medicago experiment"
author: "Ryan Farquharson"
date: "3 October 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r install packages, echo=FALSE, results='hide'}

#install.packages("ggplot2")
#install.packages("tidyverse")

library(ggplot2)
library(tidyverse)

```
## Introduction
Group B herbicides inhibit the production of branched chain amino acids, a pathway that exists in plants and also bacteria.  The sulfonylurea class of group B herbicides are generally applied to cereal crops, but in certain soil types under certain conditions can persist in soils and affect other crops that are grown in rotation with cereals.  

Legumes are included in crop rotations for multiple reasons, including the fixation of nitrogen.  Nitrogen fixation is a symbiotic process, whereby compatible saprophytic rhizobia infect root hairs of legumes resulting in the formation of nodules inside which nitrogen fixation and a mutual exchange of compounds can occur.  Both the host plants, and rhizobia, possess the pathway for production of branched chain amino acids, and acetolactate synthase (more correctly called acetohydroxyacid synthase) has been implicated in the formation of successful symbioses and potentially the transfer of nitrogen between symbionts.

This experiment was set up with the understanding that FEH-1 was tolerant of group B herbicides and would therefore provide a host background in which to determine whether the impact of group B herbicides on nitrogen fixation was entirely plant mediated or other mechanisms were at play.

The experiment was not designed to represent a real paddock in which residual herbicide were present in the soil.  As such, the application of herbicide directly to plant shoots of 17 day old seedlings represents a single set of circumstances from which some conclusions can be drawn, and from which further experiments can be designed.

## Aims
This experiment was set up to:

1) compare the nodulation and nitrogen fixation of medicago truncatula cv Herald to a near isogenic herbicide tolerant variety in the absence and presence of chlorsulfuron - a group B herbicide; and 

2) provide mechanistic insights into how group B herbicides affect nitrogen fixation.

Different stakeholders should be able to answer different questions from the results of this experiment.

Farmers and agronomists:

"Is nitrogen fixation by legumes being impacted upon when group B herbicides have been used?"
"Is it just because biomass is knocked back, or is nfix being impacted more than plant growth?"
"How well does this new variety, FEH-1 (Angel), grow and fix nitrogen in the presence and absence of herbicide?"
"How can I tell what is happening in my paddocks?
  Do I have residual group B herbicides?
  Will my plants nodulate?
  If they nodulate, will they still fix nitrogen?"

Scientists:
"What are the mechanisms by which group B herbicides impact on nitrogen fixation?"
  Plant only
    Root growth
    Root hairs
  Rhizobia only
    Saprophytic growth
  Symbiosis - Plant and rhizobia
    Signalling and nodule initiation
    Nodule development
    Mutual exchange of compounds

## Methonds in brief

Rhizobia were cultured in vitro with and without chlorsulfuron.

Plants were grown in pots containing a sand-vermiculite mix with limiting nitrogen.

Plants were inoculated or uninoculated with and without additional nitrogen.

Chlorsulfuron was applied using a spray rig.

Plants were harvested, acetylene reduction assays were performed, nodules counted and collected, biomass and nitrogen contents measured.

## The data
    
### Response variables

Since this experiment was set up with multiple objectives, a range of response variables were measured:
  Biomass of shoots, roots and nodules
  Nitrogen contents of shoots and roots plus nodules
  Pink nodule number
  Pink nodule mass
  Non-pink nodule number
  Non-pink nodule mass
  Acetylene reduction activity

From these response variables, a number of derived variables can be calculated:
  Total biomass and nitrogen
  Nodulation (nodule number or mass on a root biomass basis (pink or total))
  Nodule activity (acetylene reduction activity per unit nodule mass or per nodule (pink or total))
  Nitrogen concentrations
  Amount of nitrogen fixed

The response variables chosen for analysis will need to be chosen depending on the specific questions that need to be answered.


### Experimental factors

The experiment can be considered to be factorial with 3 factors and was blocked.

**Variety** (2 levels):  
* Herald: Medicago truncatula cv Herald;
* FEH-1: Medicago truncatula cv FEH-1.

**Herbicide** application (2 levels): 
* Herbicide: chlorsulfuron applied to 17 day old seedlings; 
* NoHerbicide: no herbicide applied applied.

**Inoculation** (5 levels): 
* O-N: uninoculated without nitrogen; 
* O+N: uninoculated with nitrogen; 
* RRI128: rhizobia strain for medicago truncatula; 
* herb: RRI128 plus herbicide in the culture; 
* washed: RRI128 plus herbicide in the culture then washed

**Block** (5 blocks):  
The experiment was set up, arranged in the glasshouse, and analysed in blocks numbered 1 to 5. Each block contained all treatments, and all pots were randomised within each block.

### How many experimental conditions?

The experiment was set up with multiple objectives, thus subsets of the data should be analysed depending on the specific research question to be answered.

The Herbicide and Variety factors can be considered as a 2 by 2 factorial (4 combinations)

Depending on which response variable is being analysed, a subset of rhizobia treatments may be considered.  
For example, it makes no sense to include the uninoculated treatments in any analysis of nodulation or nodule activity.  Subsets of the data were created accordingly.

The Block factor (5 blocks) can be considered to be uniform across all other treatments since all pots were randomised within a block.  Hence block can be considered to be an additive factor.


### Raw data

```{r read in data, echo = FALSE}
# read in raw data from csv file with headers

allsamples <- read.csv("../data/herbicide_raw_data.csv")

allsamples$Block <- factor(allsamples$Block)
allsamples$Inoculation <- factor(allsamples$Inoculation, c("RRI128", "washed", "herb", "O-N", "O+N"))
allsamples$Herbicide <- factor(allsamples$Herbicide, c("NoHerbicide", "Herbicide"))


# return the structure of the dataframe
str(allsamples)

# filer data frame to include only pots that were innoculated 

inoculated <- filter(allsamples, Inoculation %in% c("RRI128", "herb", "washed"))

uninoculated <- filter(allsamples, Inoculation %in% c("O-N", "O+N"))

str(inoculated)

```

In R it is easy to do visualisation and analysis on calculations or transformations of the raw data.

For example, instead of making a new column called totalN for the sum of shootN and bgN, simply do plots and statistical modelling of shootN + bgN.

## Visualisation

Visualisation can be useful to quickly examine the data in relation to treatments, combinations of treatments, and to other data.  The goal is to identify potential stories that we can explore further or find support for using statistical models.

### Did the experiment work?

Unfortunately, there was some nitrogen in the potting mixed used, but the uninoculated controls with and without nitrogen, we can tell that in this experiment plants were nitrogen limited.  Looking at the relationship between plant biomass and plant nitrogen by inoculation type, you can clearly see that inoculated plants lie between the uninoculated plants with and without added nitrogen.

```{r N limited, echo=FALSE}
ggplot(allsamples, aes(shootDM + bgDM, shootN + bgN, colour = Inoculation)) +
  geom_smooth(aes(fill = Inoculation), alpha = 0.2) +
  geom_point(size = 3, alpha = 0.5) +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))
```

Also, none of the uninoculated plants had nodules and there was no acetylene reduction activity in any of the uninoculated plants that were tested, indicating that there was no cross contamination in the experiment (data not shown).

### Can herbicides affect nitrogen fixation?

#### Plant N
The most simple way to determine whether the amount of nitrogen fixed has been affected by chlorsulfuron is to compare the nitrogen contents for inoculated plants with and without chlorsulfuron.

```{r plant N, echo=FALSE}
ggplot(inoculated, aes(Herbicide, shootN + bgN, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
  #scale_x_discrete("Herbicide", ) +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

```

Clearly herbicide application resulted in less nitrogen in Herald plants, but probably not as severe in FEH-1 plants.

Also note that in the absence of herbicide there was more nitrogen in Herald plants than FEH-1 plants.  In the presence of herbicide, the opposite was true.  There's an interaction if ever I saw one.

But is this just because biomass was lower?  Let's take a look.

```{r plant biomass, echo=FALSE}
ggplot(inoculated, aes(Herbicide, shootDM + bgDM, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))
```

Looks familiar, right?  On the surface it looks like the main story in this experiment is a biomass thing, rather than a specific effect on nitrogen fixation.  But we'll take a closer look to be sure.


#### Acetylene Reduction Assay
To zero in on nitrogen fixation itself, we have a measure on nitrogenase activity at the time of harvest - the acetylene reduction assay, whereby acetylene is reduced to ethylene (both of which we can measure in a GC with an FID) instead of nitrogen being reduced to ammonia.

```{r ARA, echo=FALSE}
ggplot(inoculated, aes(Herbicide, ara, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
    theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))
```

Ok, the data is noisy, but you can see that nitrogenase activity of FEH-1 and Herald plants in the absence of herbicide are on par.  Both take a hit with herbicide, but FEH-1 hangs on whilst Herald really gets knocked hard.

#### Nodulation

We spent a lot of time plucking and counting nodules. Noting that herbicide was applied to shoots later than nodule initiation time, we would expect differential effects of herbicide on pink and non-pink nodules and also nodule mass.  Given the timing of herbicide application, we would expect nodule mass to be affected more so than nodule number.

```{r nodulation plots, echo=FALSE}

ggplot(inoculated, aes(Herbicide, pinknods, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
    theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))


ggplot(inoculated, aes(Herbicide, pinknodmass, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
    theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))


ggplot(inoculated, aes(pinknods, pinknodmass, colour = Herbicide)) +
  #try adding an abline or similar +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap("Variety") +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))
```

Noting that some of the nodule initiation and development would have occurred prior to herbicide application, can we see anything interesting in the relationships between ARA and nodule number or mass?

```{r ARA and nodulation, echo=FALSE}
ggplot(inoculated, aes(pinknods, ara, colour = Herbicide)) +
  #geom_smooth(aes(fill = Inoculation)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap("Variety") +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

ggplot(inoculated, aes(pinknodmass, ara, colour = Herbicide)) +
  #geom_smooth(aes(fill = Inoculation)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap("Variety") +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))


```

Nodule number and mass are clustering strongly with herbicide treatment for Herald, but in FEH-1 we get a bit more of a continuum.

It is also interesting that there is a good relationship between plant biomass and ARA for FEH-1, but this breaks down a bit for Herald.

```{r ARA and biomass, echo=FALSE}

ggplot(inoculated, aes(bgDM + shootDM, ara,  colour = Herbicide)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap("Variety") +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

```


### What are the mechanisms
Is this purely a plant growth issue? 

Let's not forget the uninoculated controls...

```{r uninoculated plants, echo=FALSE}

ggplot(uninoculated, aes(Herbicide, shootN + bgN, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
  facet_wrap("Inoculation") +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

ggplot(uninoculated, aes(Herbicide, shootDM + bgDM, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
  facet_wrap("Inoculation") +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

```

Obviously this isn't purely a nitrogen fixation issue.  Plant growth and nitrogen is being reduced by chlorsulfuron even when uninoculated plants are supplied with N.  And interestingly, when looking only at uninoculated plants with excess nitrogen in the soil, we don't see a discrepancy between FEH-1 and Herald in the absence of herbicide.

### What about the rhizobia?
Prior to inoculation, rhizobia were cultured with (herb) and without herbicide (RRI128).  And to check whether herbicide was carrying over with the inoculum, we also did a rinsing procedure (washed).

There were not any obvious main effects, but it will be interesting to see whether the stats reveal any subtle interactions.

### Is FEH-1 a solution?

It looks like the answer is "Yes".  And "No".

You'll have noted the interactions whereby in the presence of herbicide, FEH-1 generally performed better than Herald.  But in the absence of herbicide, Herald generally performed better that FEH-1.

So if group B herbicides are used in-crop, or are have been used in preceding crops and are expected to still reside in the soil, FEH-1 would be the best choice.  If, group B herbicides are not an issue, stick with Herald.


## Now for some stats

We've looked at the data.  We've seen some stories.  Do the stats support them?  Is there anything we've missed?

Let's do some statistical modelling to test these ideas...

```{r install stats packages, echo=FALSE}

#install.packages("lmerTest")
#install.packages("emmeans")
library(lmerTest)
library(emmeans)

```

First up, let's look at nitrogenase activity.

First we'll try a linear model looking for interactions but putting Block as an additive fixed effect (ie Block effect is assumed to be constant across all other treatments).

```{r ara lm}
aralm <- lm(ara~Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(aralm)
plot(aralm, which = 1)
```
Note that the residuals show heteroscedacity.  That is, as the fitted values increase, the residuals increase.  Also note that there is curvature in the relationship between fitted values and residuals.  We'll need to do some transformations, but first let's compare this model with an lmer whereby we put Block in the random effects part of the model.

```{r ara lmer}
aralmer <- lmer(ara~Variety*Herbicide*Inoculation + (1|Block), data=inoculated)
anova(aralmer)
plot(aralmer, which = 1)
```
We get the same thing.  Unfortunately, with lmer, we don't get the red line showing trends in fitted values vs residuals.  Here, we don't get information about the Block effect because by specifying it as part of the random effects we are saying that we don't care about it.  Personally, I find it reassuring that the block effect was quite strong.  All that effort that went into randomising treatments within blocks, setting up, running and analysing the experiment in blocks, was worth it.  We now have more power to look at the treatment effects.

Now for some transformations.  Noting that the residuals increase as the fitted values increase, a square root transformation should be useful.

```{r sqrt ara lm}
sqrt_ara_lm <- lm(sqrt(ara)~Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(sqrt_ara_lm)
plot(sqrt_ara_lm, which = 1)
```

This is an improvement.  It's not perfect - there may still be some structure in that we have few samples on the left, but I think it is acceptable, and we will be interpreting the data along with other evidence.

Again, we can repeat the analysis using lmer and put the block effect in with the random effects and we get the same thing.

```{r sqrt ara lmer}
sqrt_ara_lmer <- lmer(sqrt(ara)~Variety*Herbicide*Inoculation + (1|Block), data=inoculated)
anova(sqrt_ara_lmer)
#summary(sqrt_ara_lmer)
#emmeans(sqrt_ara_lmer, ~Variety*Herbicide, type = "response")
plot(sqrt_ara_lmer)

```

Out of interest, we can try a log transformation. Using lm we get 4 plots.  I'll need to do some digging to understand what they all mean.

```{r log ara lm}
log_ara_lm <- lm(log(ara+1)~Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(log_ara_lm)
#summary(log_ara_lm)
#emmeans(log_ara_lm, ~Variety*Herbicide, type = "response")
plot(log_ara_lm, which=1)

```

```{r log ara lmer}
log_ara_lmer <- lmer(log(ara+1)~Variety*Herbicide*Inoculation + (1|Block), data=inoculated)
anova(log_ara_lmer)
#summary(log_ara_lmer)
#emmeans(log_ara_lmer, ~Variety*Herbicide, type = "response")
plot(log_ara_lmer)
```

Looking at these statistical models, they are all telling us a similar story.  The rhizobia treatments make little difference, and there is an interaction between Variety and Herbicide.  

So, what does the interaction mean?  Well, the effect of Herbicide is dependent on Variety, or the effect of Variety is dependent on Herbicide.

We can use emmeans to give us mean values and contrasts.

```{r}
emmeans(log_ara_lm, ~Variety*Herbicide, type = "response")
emmeans(log_ara_lmer, ~Variety*Herbicide, type = "response")
```
Using the lm or lmer, we get the same response but the SE, df and confidence intervals are different because of the way we have averaged, with and without the block effect I suspect.

The story is still the same, though.  ARA of FEH-1 and Herald are the same in the absence of herbicide.  But in the presence of herbicide, ARA of Herald drops more than FEH-1.


So that was a fairly in depth look at statistical modelling of ARA.  There is still a lot more to look at.  I am going to cut straight to what I think are the most appropriate statistical models for the variables of interest, and use the stats to support or otherwise the interpretations of the data, whilst being on the lookout for any unanticipated findings. 


### Plant nitrogen
Given that our main question is whether herbicides affect the amount of nitrogen fixed by medicago, and whether using a herbicide tolerant variety can be a solution to this problem, let's look at the amount of nitrogen in the plants.  We will look at inoculated plants only (not the O-N or O+N treatments).

```{r Plant nitrogen statistical model I}
TotalPlantN_lm <- lm(shootN + bgN ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(TotalPlantN_lm)
plot(TotalPlantN_lm, which = 1)
```

```{r Plant nitrogen statistical model II}
sqrtTotalPlantN_lm <- lm(sqrt(shootN + bgN) ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(sqrtTotalPlantN_lm)
plot(sqrtTotalPlantN_lm, which = 1)
```

Taking the square root improves the residuals plot a bit, but doesn't change the story.  There is a very strong Variety by Herbicide interaction, and a hint of a Variety by Inoculation interaction with the sqrt transformed data.

```{r total plant N means}
emmeans(TotalPlantN_lm, ~ Variety*Herbicide)
```

This is a very clear example of an interaction.  In the absence of herbicide, Herald plants contained more nitrogen than FEH-1 plants.  In the presence of herbicide, nitrogen contents declined more for Herald plants than FEH-1 plants so that FEH-1 plants had more nitrogen the Herald plants.  We already knew that from the visualisation.  The stats back it up.

Just to satisfy myself, I will repeat using lmer and put the Block effect into the noise.

```{r sqrtTotalPlantN_lmer}
sqrtTotalPlantN_lmer <- lmer(sqrt(shootN + bgN) ~ Variety*Herbicide*Inoculation + (1|Block), data=inoculated)
anova(sqrtTotalPlantN_lmer)
plot(sqrtTotalPlantN_lmer)
emmeans(sqrtTotalPlantN_lmer, ~ Variety*Herbicide, type="response")
```

```{r TotalPlantN_lmer}
TotalPlantN_lmer <- lmer(shootN + bgN ~ Variety*Herbicide*Inoculation + (1|Block), data=inoculated)
anova(TotalPlantN_lmer)
plot(TotalPlantN_lmer)
emmeans(TotalPlantN_lmer, ~ Variety*Herbicide)
```
Different confidence intervals, but same story really.
 
What if we look at shoots and below ground separately?

First the shoots...

```{r Shoot nitrogen statistical model I}
shootN_lm <- lm(shootN ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(shootN_lm)
plot(shootN_lm, which =1)
emmeans(shootN_lm, ~ Variety*Herbicide)
```

The shoots are responding with that classic Variety by Herbicide interaction.

Now for the roots+nodules...

```{r below ground nitrogen statistical model I}
bgN_lm <- lm(bgN ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(bgN_lm)
plot(bgN_lm)
emmeans(bgN_lm, ~ Variety*Herbicide)
emmeans(bgN_lm, pairwise ~ Herbicide)
emmeans(bgN_lm, pairwise ~ Variety)
```
A hint of an interaction, but we can only be confident about the main effects of Variety and Herbicide, where NoHerbicide was greater than Herbicide, and Herald was greater than FEH-1.

This is interesting.  In our visualisation we only looked at whole plant nitrogen and biomass.  But the statistical modelling suggests that the shoots and roots+nodules are responding differently. This is poignant given that the herbicide was applied to the shoots, and nodulation occurs in the roots.  This is useful information that could inform the design of a subsequent experiment.

Time permitting it would be good to be able to plot shoots and bg on the same plot, above and below the x axis.

### Plant biomass

#### Shoots
```{r shoot DM}
shootDM_lm <- lm(shootDM ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(shootDM_lm)
plot(shootDM_lm)
emmeans(shootDM_lm, ~ Variety*Herbicide)
```

```{r sqrt shoot DM}
sqrtshootDM_lm <- lm(sqrt(shootDM) ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(sqrtshootDM_lm)
plot(sqrtshootDM_lm)
emmeans(sqrtshootDM_lm, ~ Variety*Herbicide, type="response")
emmeans(sqrtshootDM_lm, ~ Herbicide*Inoculation, type="response")
```

I log transformed the data which did not make much difference to the residual plot.  The sqrt transformation looked good.  We have a strong Herbicide by Variety interaction again.  We also have a Variety by Inoculation interaction because the 'washed' treatment without herbicide applied to the plant, had higher shoot DM than all others.  This result is difficult to interpret.

#### Roots
```{r root DM}
bgDM_lm <- lm(bgDM ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(bgDM_lm)
plot(bgDM_lm)
emmeans(bgDM_lm, ~ Variety*Herbicide, type="response")
```
```{r sqrt root+nodule DM}
sqrtbgDM_lm <- lm(sqrt(bgDM) ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(sqrtbgDM_lm)
plot(sqrtbgDM_lm)
emmeans(sqrtbgDM_lm, ~ Variety*Herbicide, type="response")
```

The sqrt transformation improved the homoscedacity of the residuals, and strengthened the Variety by Herbicide interaction.  For root+nodule DM, in the absence of herbicide FEH-1 had higher bgDM.  In the presence of herbicide, bgDM declined for both FEH-1 and Herald, more so for Herald, and then both ended up with the same bgDM.

### Nodulation
Let's look at nodules, which are the organs inside which the symbiosis happens.  To form nodules, there is some complex signalling between symbionts, infection of roots and development of nodules.  Then there is an exchange of compounds.  In reality, this experiment is looking at a very specific set of circumstances in which herbicide was applied to the shoots some time after much infection would have occurred.  We have nodule numbers which are somewhat arbitrary because clusters of nodules can form, and not all non-pink nodules may have been identified because they may have been too small to notice. We also have nodule masses, which are likely to be more reliable, particularly for pink nodules.

```{r nodulation stats}
sqrtpinknodmass_lm <- lm(sqrt(pinknodmass) ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(sqrtpinknodmass_lm)
plot(sqrtpinknodmass_lm)
emmeans(sqrtpinknodmass_lm, ~ Variety*Herbicide, type="response")
emmeans(sqrtpinknodmass_lm, ~ Herbicide*Inoculation, type="response")
emmeans(sqrtpinknodmass_lm, ~ Variety*Inoculation, type="response")
```

This is a case where the stats are flagging some potentially interesting interactions that are easy to miss during visualisation due to noisy data.  Although it would be easy to discount these interactions as unimportant (and they are marginal in this experiment), we should try to interpret them in terms of their potential implications in a real world situation where a precautionary approach may be advisable.  

We have seen differential effects of herbicide application to shoots on shoot and below ground biomass and nitrogen.  We are now seeing some different responses when herbicide has been added to flasks during culturing of the rhizobia, compared to no herbicide added to the flask, or the pellet of rhizobia washed, but only for FEH-1.  This is a one off observation at a more marginal p level so a subsequent experiment designed to examine the topical effects of herbicide on nodule development in the different plant genetic backgrounds would be advisable before drawing conclusions, especially given the fact that residual herbicides are of particular concern for many farmers.

Now back to the Variety by Herbicide interaction which we have more confidence in... it is more similar to the shoot N and biomass interaction whereby in the absence of herbicide, Herald had greater pink nodule mass than FEH-1, whereas in the presence of herbicide, the pink nodule mass of Herald decline more than that of FEH-1, ending up lower and close to zero.

SO we know that shoot and root biomass are affected.  Is nodule mass more or less sensitive that root biomass?  TO find out, we need to analyse nodule mass divided by root mass.


```{r sqrt nodule DM per root DM}
sqrtnodDMperrootDM_lm <- lm(sqrt(pinknodmass / rootDM) ~ Variety*Herbicide*Inoculation + Block, data=inoculated)
anova(sqrtnodDMperrootDM_lm)
plot(sqrtnodDMperrootDM_lm)
emmeans(sqrtnodDMperrootDM_lm, ~ Variety*Herbicide, type="response")
emmeans(sqrtnodDMperrootDM_lm, ~ Variety*Inoculation, type="response")
```
Again, there are some interesting interactions here that deserve some follow up.  But the clear message is that nodule mass is more sensitive to herbicide than root mass, and it depends on the Variety.  

### Uninoculated plants

We have focused on inoculated plants so far.  Let's take a closer look at the uninoculated plants which were included as controls.  

```{r uninoculated N}
uninoculatedN_lm <- lm(shootN + bgN ~ Variety*Herbicide*Inoculation + Block, data=uninoculated)
anova(uninoculatedN_lm)
plot(uninoculatedN_lm, which = 1)
emmeans(uninoculatedN_lm, ~ Variety*Herbicide)
emmeans(uninoculatedN_lm, ~ Herbicide*Inoculation)
```
The residuals don't look pretty, and that's because the treatment combinations were so different from each other.  If I subset the data and look only at the O+N treatment...

```{r plusN N}
plusN <- filter(allsamples, Inoculation %in% c("O+N"))

plusNN_lm <- lm(shootN + bgN ~ Variety*Herbicide + Block, data=plusN)
anova(plusNN_lm)
plot(plusNN_lm, which = 1)
emmeans(plusNN_lm, ~ Herbicide)
```
There was a clear main effect of herbicide on plant N.

```{r uninoculated DM}
uninoculatedDM_lm <- lm(shootDM + bgDM ~ Variety*Herbicide*Inoculation + Block, data=uninoculated)
anova(uninoculatedDM_lm)
plot(uninoculatedDM_lm, which = 1)
emmeans(uninoculatedDM_lm, ~ Variety*Herbicide)
emmeans(uninoculatedDM_lm, ~ Herbicide*Inoculation)
```

Could do with a transformation here, but the results are quite clear. The biomass response is a bit different from the N response.  For O-N plants, the nitrogen content didn't really drop with herbicide application.  However dry matter did decline, due in part to timing of application and continued growth until harvest in plants without herbicide accompanied by a decline in N concentration.

## Bringing it all together

Overall, the statistical modelling support the stories we discovered during the visualisation.  In addition the statistical modelling also revealed some subtleties that we overlooked.

The key points are that:

* FEH-1 is not fully tolerant to group B herbicides. However in the presence of herbicide, FEH-1 can continue to grown and fix nitrogen whereas Herald was knocked out.
* Symbiotic nitrogen fixation is not as optimal in FEH-1 as in Herald in the absence of herbicide, which is also reflected in growth.
* It appears that there are similar treatment effects of Variety and Herbicide on nitrogen contents and biomass, but there is some plasticity there.
* Shoots were affected differently to roots, which may be due to the topical application of herbicide to the shoots in this experiment.
* Despite the application of herbicide to the shoots, nodule mass was more sensitive to herbicide than root mass.  This may simply be a matter of timing of herbicide application.
* There were some interactions involving rhizobia treatments indicating potential carryover of herbicide from cultures of rhizobia and also the potential importance of topical exposure to herbicide.
* This experiment was a an attempt to use a herbicide tolerant host to identify potential mechanisms by which group B herbicides affect nitrogen fixation.  It was a good first look and much was learnt about the experimental system.  However some well designed experiments are needed to better represent real agricultural systems, and also to get a better mechanistic understanding of how group B herbicides affect nodulation and nitrogen fixation.


## What did I learn?

### Approach to experiments and analysis
This experiment was done more than 10 years ago.  It was interesting to compare my approach this time to my approach back then.
I deliberately started from scratch, cutting the data right back to the raw data.  In going through the exploration and analysis I had to reacquaint myself with the questions I originally set out to answer. This made it really obvious that experimental design should be done to answer specific questions, then the statistical modelling flows very easily from the experimental design.

### Visualisation
Visualisation is super useful both for data exploration and data explanation.  In the past I had taken the raw data, done a heap of calculations then run through lots of statistics looking for differences.  This can soon turn into a big muddle.  Now my approach is to focus on the questions, design the experiments, use visualisations to explore the data, look for stories, then run the statistical models to test those stories and look for things I may have missed.  Finally, and I haven't gotten to this yet, work on visualisations to effectively communicate what the data is telling you to your audience.

### Statistical modelling
Make the statistical modelling work for you.  It should flow naturally from the experimental design, which should flow naturally from the questions you set out to answer.  Think about statistical modelling from the perspective of supporting your data to answer the questions you are interested in, rather than it being an exercise to get an arbitrary p value just so you can a paper past some reviewer who doesn't really understand p values.

### Pipelines and documentation

Using Rmarkdown is an excellent way to organise, conduct and document your work. It enables you to bring together the data, the exploratory visualisation, the statistical modelling, the explanatory visualisation and the reporting into a single document that is easily to update with version control, easy to use for presenting and discussing, and gives you the ability to make into a nice shiny document as an html or pdf, all in the one place. Brilliant.  

No doubt markdown has a lot of functionality that I still need to explore, such as pandoc formatting for the final document.  But even before that, you have the ability to show or hide chunks or sections, keep all the code and choose whether or not it, warning messages or results appear in the final document.  

### Next steps
I would also like to improve my ggplot skills, and think more about how to use visualisation for explanation and publication, not just exploration.
I would like to get more proficient at r markdown, to the point where I can use it to produce publication quality reports and manuscripts for submission.
A large part of my future work will be in chemometrics using multivariate analysis, principal components analysis, partial least squares regression, piece-wise direct standardisation.
I will also be doing more modelling work so sampling data from distributions and outputting model outcomes probabalistically and generating explanatory visualisations would be very useful.

### Questions
Subsetting data
Specifying which contrasts you want
What people ended up plotting for reporting purposes - when the visualisations already tell a good story, what do you add? ablines, confidence intervals..
