---
title: "Medicago experiment"
author: "Ryan Farquharson"
date: "3 October 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install packages, echo=FALSE}

install.packages("ggplot2")
install.packages("tidyverse")

library(ggplot2)
library(tidyverse)

```

## Aims
This experiment was set up to:

1) compare the nodulation and nitrogen fixation of medicago truncatula cv Herald to a near isogenic herbicide tolerant variety in the absence and presence of chlorsulfuron - a group B herbicide; and 

2) provide mechanistic insights into how group B herbicides affect nitrogen fixation.

Different stakeholders should be able to answer different questions from the results of this experiment.

Farmers and agronomists:

"Is nitrogen fixation by legumes being impacted upon when group B herbicides have been used?"
"Is it just because biomass is knocked back, or is nfix being impacted more than plant growth?"
"How well does this new variety, FEH-1 (Angel), grow and fix nitrogen in the presence and absence of herbicide?"
"How can I tell what is happening in my paddocks?
  Do I have residual group B herbicides?
  Will my plants nodulate?
  If they nodulate, will they still fix nitrogen?"

Scientists:
"What are the mechanisms by which group B herbicides impact on nitrogen fixation?"
  Plant only
    Root growth
    Root hairs
  Rhizobia only
    Saprophytic growth
  SYmbiosis - Plant and rhizobia
    Signalling and nodule intitiation
    Nodule development
    Mutual exchange of compounds

## Methonds in brief

Rhizobia were cultured in vitro with and without chlorsulfuron.

Plants were grown in pots containing a sand-vermiculite mix with limiting nitrogen.

Plants were inoculated or uninoculated with and without additional nitrogen.

Chlorsulfuron was applied using a spray rig.

Plants were harvested, acetylene reduction assays were performed, nodules counted and collected, biomass and nitrogen contents measured.

## The data
    
### Response variables

Since this experiment was set up with multiple objectives, a range of response variables were measured:
  Biomass of shoots, roots and nodules
  Nitrogen contents of shoots and roots plus nodules
  Pink nodule number
  Pink nodule mass
  Non-pink nodule number
  Non-pink nodule mass
  Acetylene reduction activity

From these response variables, a number of derived variables can be calculated:
  Total biomass and nitrogen
  Nodulation (nodule number or mass on a root biomass basis (pink or total))
  Nodule actiivty (acetylene reduction activity per unit nodule mass or per nodule (pink or total))
  Nitrogen concentrations
  Amount of nitrogen fixed

The response variables chosen for analysis will need to be chosen depending on the specific questions that need to be answered.


### Experimental factors

The experiment can be considered to be factorial with 3 factors and was blocked.

Variety (2 levels):  
  - Herald: Medicago truncatula cv Herald;
  - FEH-1: Medicago truncatula cv FEH-1.

Herbicide application (2 levels): 
  - Herbicide: chlorsulfuron applied to 17 day old seedlings; 
  - NoHerbicide: no herbicide applied applied.

Rhizobia (5 levels): 
  - 0-N: uninoculated without nitrogen; 
  - 0+N: uninoculated with nitrogen; 
  - RRI128: rhizobia strain for edicago truncatula; 
  - herb: RRI128 plus herbicide in the culture; 
  - washed: RRI128 plus herbicide in the culture then washed

Block (5 blocks):  
  The experiment was set up, arranged in the glasshouse, and analysed in blocks numberred 1 to 5. Each block contained all treatments, and all pots were randomised within each block.

### How many experimental conditions?

The experiment was set up with multiple objectives, thus subsets of the data should be analysed depending on the specific research question to be answered.

The Herbicide and Variety factors can be considered as a 2 by 2 factorial (4 combinations)

Depending on which repsonse variable is being analysed, a subset of rhizobia treatments may be considered.  
For example, it makes no sense to include the uninoculated treatments in any analysis of nodulation or nodule activity.


The Block factor (5 blocks) can be considered to be uniform across all other treatments since all pots were randomised within a block.  Hence block can be considered to be an additive factor.


### Raw data

```{r}
# read in raw data from csv file with headers

getwd()

allsamples <- read.csv("../data/herbicide_raw_data.csv")

allsamples$Block <- factor(herb$Block)
allsamples$Inoculation <- factor(herb$Inoculation, c("RRI128", "washed", "herb", "O-N", "O+N"))
allsamples$Herbicide <- factor(herb$Herbicide, c("NoHerbicide", "Herbicide"))


# return the structure of the dataframe
str(allsamples)

# filer data frame to include only pots that were innoculated 

inoculated <- filter(allsamples, Inoculation %in% c("RRI128", "herb", "washed"))

uninoculated <- filter(allsamples, Inoculation %in% c("O-N", "O+N"))

str(inoculated)

```

In R it is easy to do visualisation and analysis on calculations or transformations of the raw data.

For example, instead of making a new column called totalN for the sum of shootN and bgN, simply do plots and statistical modelling of shootN + bgN.

## Visualisation

Visualisation can be useful to quickly examine the data in relation to treatments, combinations of treatments, and to other data.  The goal is to identify potential stories that we can explore further or find support for using statistical models.

### Did the experiment work?

Unfortunately, there was some nitrogen in the potting mixed used, but the unoninoculated controls with and without nitrogen, we can tell that in this experiment plants were nitrogen limited.  Looking at the relationship between plant biomass and plant nitrogen by inoculation type, you can clearly see that inoculated plants lie between the uninoculated plants with and without added nitrogen.

```{r}
ggplot(allsamples, aes(shootDM + bgDM, shootN + bgN, colour = Inoculation)) +
  geom_smooth(aes(fill = Inoculation)) +
  geom_point(size = 3, alpha = 0.5) +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))
```

Also, none of the uninoculated plants had nodules and there was no acetylene reduction activity in any of the uninoculated plants that were tested, indicating that there was no cross contamination in the experiment (data not shown).

### Can herbicides affect nitrogen fixation?

#### Plant N
The most simple way to determine whether the amount of nitrogen fixed has been affected by chlorsulfuron is to compare the nitrogen contents for inoculated plants with and without chlorsulfuron.

```{r}
ggplot(inoculated, aes(Herbicide, shootN + bgN, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
  #scale_x_discrete("Herbicide", ) +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

```

Clearly herbicide application resulted in less nitrogen in Herald plants, but probably not as severe in FEH-1 plants.

Also note that in the absence of herbicide there was more nitrogen in Herald plants than FEH-1 plants.  In the presence of herbicide, the opposite was true.  There's an interaction if ever I saw one.

But is this just because biomass was lower?  Let's take a look.

```{r}
ggplot(inoculated, aes(Herbicide, shootDM + bgDM, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))
```

Looks familiar, right?  Lools like this might just be a biomass thing, rather than a specific effect on nitrogen fixation.  But we'll take a closer look to be sure.


#### ARA
To zero in on nitrogen fixation itself, we have a measure on nitrogenase activity at the time of harvest - the acetylene reduction assay, whereby acetylene is reduced to ethylene (both of which we can measure in a GC with an FID) instead of nitrogen being reduced to ammonia.

```{r}
ggplot(inoculated, aes(Herbicide, ara, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
    theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))
```

Ok, the data is noisy, but you can see that nitrogenase activity of FEH-1 and Herald plants in the absence of herbicide are on par.  Both take a hit with hebricide, but Herald really gets knocked hard.

#### Nodulation

We spent a lot of time plucking and counting nodules. Noting that herbicide was applied to shoots later than nodule initiation time, we would expect differential effects of herbicide on pink and non-pink nodules and also nodule mass.  Given the timing of herbicide application, we would expect nodule mass to be affected more so than nodule number.

```{r nodulation}

ggplot(inoculated, aes(Herbicide, pinknods, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
    theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))


ggplot(inoculated, aes(Herbicide, pinknodmass, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
    theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))


ggplot(inoculated, aes(pinknods, pinknodmass, colour = Herbicide)) +
  #geom_smooth(aes(fill = Inoculation)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap("Variety") +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))
```


Noting that some of the nodule initiation and development would have occured prior to herbicide application, can we see anything interesting in the realtionships between ARA and nodule number or mass?

```{r}
ggplot(inoculated, aes(pinknods, ara, colour = Herbicide)) +
  #geom_smooth(aes(fill = Inoculation)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap("Variety") +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

ggplot(inoculated, aes(pinknodmass, ara, colour = Herbicide)) +
  #geom_smooth(aes(fill = Inoculation)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap("Variety") +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))


```

Nodule number and mass are clustering strongly with herbicide treatment for Herald, but in FEH-1 we get a bit more of a continuum.

It is also interesting that there is a good relationship between plant biomass and ARA for FEH-1, but this breaks down for Herald.

```{r}

ggplot(inoculated, aes(bgDM + shootDM, ara,  colour = Herbicide)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap("Variety") +
  theme(panel.grid = element_blank(),
         panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

```



#### What can other relationships tell us?

What's driving what?  Given that there is a difference in the tolerance of the two varieties to herbicide, can we can use this difference to tell us whether reduced growth and biomass is affecting nitrogen fixation, or vice versa?


### What are the mechanisms
Is this purely a plant growth issue? 

Let's not forge the uninocuated controls...

```{r}

ggplot(uninoculated, aes(Herbicide, shootN + bgN, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
  facet_wrap("Inoculation") +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

ggplot(uninoculated, aes(Herbicide, shootDM + bgDM, colour = Variety)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75), size = 4, alpha = 0.5) +
  facet_wrap("Inoculation") +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

```

Obviously this isn't purely a nitrogen fixation issue.  Plant growth and nitrogen is being reduced by chlorsulfuron even when uninoculated plants are supplied with N.  

### What about the rhizobia?
Prior to inoculation, rhizobia were cultured with (herb) and without herbicide (RRI128).  And to check whether herbicide was carrying over with the inoculum, we also did a rinsing proceedure (washed).

There were not any obvious main effects, but it will be interesting to see whether the stats reveal any subtle interactions.

### Is FEH-1 a solution?

Yes.  And no.

You'll have noted the interactions whereby in the presence of herbicide, FEH-1 generally performed better than Herald.  But in the absence of herbicde, Herald generally performed better that FEH-1.

So if group B herbicides are used in-crop, or are have been used in preceeding crops and are expected to still reside in the soil, FEH-1 would be the best choice.  If, group B herbicides are not an issue, stick with Herald.


## To summarise



## Now for some stats

We've looked at the data.  We've seen some stories.  Do the stats support them?  Is there anything we've missed?

Let's do some statistical modelling to test these ideas...

```{r install stats packages, echo=FALSE}

install.packages("lmerTest")
install.packages("emmeans")
library(lmerTest)
library(emmeans)


```



```{r}
aralmer <- lmer(ara~Variety*Herbicide*Inoculation + (1|Block), data=inoculated)
anova(aralmer)
summary(aralmer)
emmeans(aralmer, ~Variety*Herbicide, type = "response")
plot(aralmer, which = 1)
```

```{r}
sqrt_ara_lmer <- lmer(sqrt(ara)~Variety*Herbicide*Inoculation + (1|Block), data=inoculated)
anova(sqrt_ara_lmer)
summary(sqrt_ara_lmer)
emmeans(sqrt_ara_lmer, ~Variety*Herbicide, type = "response")
plot(sqrt_ara_lmer)


```


```{r}
log_ara_lmer <- lmer(log(ara+1)~Variety*Herbicide*Inoculation + (1|Block), data=inoculated)
anova(log_ara_lmer)
summary(log_ara_lmer)
emmeans(log_ara_lmer, ~Variety*Herbicide, type = "response")
plot(log_ara_lmer)

View(inoculated)

```


## Questions for Terry

ELegant ways to subset (levels)

Transformation - e.g. ara (square root?)

Getting specific contrasts and showing that the differences are different from each other

Derived or calculated variables e.g. amount of nitrogen fixed, fixed per nodule, fixed per unit nodule mass

Nested designs - e.g. innoculated/uninoculated +-N, innoculated +H, -H, washed




